---
title: "ABA as a framework for reinforcement-learning problems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ABA as a framework for reinforcement-learning problems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(operantlunar)
```

A reinforcement-learning problem is a reinforcement problem: an agent's
behaviour is shaped by the reward contingencies of its environment. Applied
behaviour analysis is the science of analysing and changing behaviour through
reinforcement contingencies. So the ABA toolkit ([`aba_toolkit()`]) is an
alternative analytic framework for the problems posed by a Gymnasium
environment: its instruments diagnose what maintains an agent's behaviour and
change it by manipulating the contingency, rather than by reading weights or
gradients.

The bridge is the *contingency*. A Gymnasium environment exposes
`reset`/`step`; [`contingency_env()`] wraps any environment whose `step` returns
a named `channels` vector and recomputes reward as the sum of the **active**
channels. `set_active()` is therefore the single lever every ABA instrument
pulls. [`as_channel_env()`] turns an ordinary single-reward gym environment
(e.g. [`make_gym()`]) into a one-channel environment, so the same instruments
apply to off-the-shelf tasks.

## The mapping

```{r mapping}
aba_gym_mapping()
```

## Functional analysis = reward-channel ablation

The applied question "what reinforcer maintains this behaviour?" becomes the RL
question "what reward component drives this policy?". [`gym_functional_analysis()`]
trains the agent under each single-channel condition and a no-channel play
control, then identifies the maintaining channel as the condition whose target
response rate is elevated over control. It recovers a planted function:

```{r fa, eval = FALSE}
fa <- gym_functional_analysis(true_function = "escape")
fa$identified_channel
#> "escape"
plot_gym_functional_analysis(fa)
```

Verified result: the matching condition shows a target-response rate near 0.98
while every other condition and the play control sit near 0.03, and
`identified_channel == true_function` for all of attention, escape, tangible,
and goal. Ablating the channels isolates the one the behaviour depends on.

## Extinction = withdraw the maintaining channel

Withholding the maintaining reinforcer should weaken the behaviour.
[`gym_extinction()`] acquires a policy with the channel active, then withdraws it
(channel inactive) while continuing to train, probing retained behaviour by
evaluating with the channel restored. On the real `FrozenLake-v1` task:

```{r ext, eval = FALSE}
builder <- function() as_channel_env(make_gym("FrozenLake-v1", is_slippery = FALSE), "goal")
gym_extinction(builder)
```

Verified result: the acquired greedy policy reaches the goal on 100% of probes,
then collapses to 0% within roughly ten unreinforced episodes. Extinction under
a deterministic greedy value learner is abrupt — once the path's value decays
below the alternatives, the policy is abandoned wholesale — which is itself an
honest reading the framework surfaces.

## Differential reinforcement = reallocate the contingency

The function-based intervention "reinforce an alternative while the problem
behaviour is on extinction" becomes contingency-switching on a navigation task.
[`gym_dra()`] reinforces a problem arm (baseline), then switches the active
channel so the problem arm is extinguished and an alternative arm is reinforced:

```{r dra, eval = FALSE}
demo <- gym_dra()
plot_gym_dra(demo)
```

Verified result: the agent reaches the problem arm on ~0.84 of baseline episodes
and reallocates to the alternative arm at ~0.98 by the end of treatment. The
contingency engine that underlies DRA/FCT transfers directly to a multi-state
gym environment.

## What this is and is not

These instruments validate the ABA *logic* on the Gymnasium contract with
reinforcement-driven agents; they are not clinical procedures, and functional
analysis remains one scoped instrument rather than a prerequisite. What the
framework adds to an RL problem is a vocabulary of contingency manipulations —
ablate a channel to attribute a behaviour, withdraw it to test persistence,
switch it to redirect — that read and change behaviour in the environment's own
terms.
