---
title: "An operant-conditioning primer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An operant-conditioning primer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(operantlunar)
```

This primer orients a newcomer to the operant-conditioning vocabulary that
`operantlunar` is built on. Every definition below is drawn from
`operant_glossary()` and was checked against the literature before entry; the
sources are the bundled handbook corpus plus the original papers, listed in
`operant_bibliography()`.

## The one idea

Operant conditioning studies how the **consequences** of a voluntary behaviour
change its future probability. Thorndike's *law of effect* (1898, 1911) stated
the seed: responses followed by satisfaction are stamped in, those followed by
discomfort stamped out. Skinner (1938) made it functional — replacing
"satisfaction" with **reinforcement** (anything that increases a behaviour) and
"discomfort" with **punishment** (anything that decreases it), each defined
purely by its effect. The unit of analysis is the **three-term contingency**:
an antecedent signals that a behaviour will produce a consequence.

The same logic is what reinforcement learning later formalised, which is why
this package can compare *reward maximisation* (the law of effect taken to its
optimum) against *melioration* (the rule behind the matching law) using one
learning skeleton and swapping only the update rule.

## How the vocabulary maps onto this package

```{r map, echo = FALSE, results = "asis"}
map <- tibble::tribble(
  ~`Operant concept`, ~`In operantlunar`,
  "Reinforcement schedule (FR/VR/FI/VI)", "`make_schedule()`, `operant_chamber()`",
  "Concurrent schedules / choice", "`concurrent_schedule()`, `schedule_matching_table()`",
  "Matching law", "`fit_matching_general()`, `melioration_agent()` selection",
  "Melioration vs maximisation", "`melioration_trap()`, `differentiation_matrix()`",
  "Extinction", "`extinction_experiment()`",
  "Delay / self-control", "`self_control_env()`, `self_control_experiment()`",
  "Differential reinforcement of low rates", "`drl_chamber()`, `drl_experiment()`",
  "Reinforcer devaluation (habit vs goal)", "`devaluation_experiment()`"
)
knitr::kable(map)
```

## Glossary

```{r glossary, echo = FALSE, results = "asis"}
g <- operant_glossary()
for (cat in unique(g$category)) {
  cat("\n### ", cat, "\n\n", sep = "")
  sub <- g[g$category == cat, c("term", "definition", "primary_source")]
  print(knitr::kable(sub, col.names = c("Term", "Definition", "Primary source"), row.names = FALSE))
  cat("\n")
}
```

## Bibliography

The handbook corpus is held in GCS at
`gs://sondre_brreg_data/losore_schema_engine/aba_references/source_pdfs/`; the
primary sources are the original papers the definitions rest on.

```{r biblio, echo = FALSE, results = "asis"}
b <- operant_bibliography()
knitr::kable(b[, c("key", "type", "citation")], col.names = c("Key", "Type", "Citation"), row.names = FALSE)
```
