---
title: "Differentiating reward maximization from melioration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differentiating reward maximization from melioration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r lib}
library(operantlunar)
```

## The question

Two accounts describe how choice tracks reinforcement. The **law of effect**
says behavior moves toward the action with the highest long-run return: reward
maximization. **Melioration** (the mechanism behind the matching law) says
behavior moves toward the action with the higher *local* return, evaluated
myopically. On many environments the two coincide. They come apart only when
choosing an option changes the local returns, or when a smaller-sooner reward
competes with a larger-later one.

`operantlunar` holds the learning skeleton fixed and swaps the update rule, then
runs each rule through operant paradigms designed to separate the two accounts.

## The rule library

Every rule plugs into every paradigm through a registry:

```{r registry}
names(agent_registry())
sapply(c("q_learning", "melioration", "melioration_rate", "win_stay_lose_shift"), agent_kind)
```

```{r make-agent, eval = FALSE}
ag <- make_agent("double_q", n_actions = 2L, horizon = 30000L)
```

## The melioration trap

The trap is the canonical dissociation: option A pays more, but the more A is
chosen the worse both options become. The rate-maximizing allocation differs
from the matching allocation.

```{r trap-analytic}
tr <- melioration_trap()
tr$optimum()
tr$matching_point()
```

A reward maximizer settles near the optimum; gradient-bandit melioration is
pulled to the matching point and earns less. (Full run elided for build speed.)

```{r trap-exp, eval = FALSE}
melioration_trap_experiment(n_steps = 60000)
```

## More paradigms

```{r paradigms-code, eval = FALSE}
prob_matching_experiment(probs = c(0.75, 0.25))   # matching vs exclusive choice
self_control_experiment()                          # delay discounting
drl_experiment()                                   # low-rate responding
devaluation_experiment()                           # habit vs goal-directed
```

Probability matching is produced by the rate-tracking rule, not by the
gradient-bandit melioration (which maximizes on a stationary bandit). Self-control
and reinforcer devaluation cleanly separate myopic from far-sighted control.

## The differentiation matrix

The capstone scores each rule on each paradigm in `[0, 1]`, where 1 is
reward-maximizing and lower values indicate matching or suboptimal behavior. A
small two-rule, two-paradigm version runs here; the full instrument uses more
rules, more paradigms, and a larger step budget.

```{r matrix, fig.width = 5, fig.height = 3}
dm <- differentiation_matrix(
  rules = c("q_learning", "melioration"),
  paradigms = c("prob_matching", "trap"),
  n_steps = 5000
)
dm$wide
plot_differentiation_matrix(dm)
```

## Function approximation and LunarLander

Tabular state binning is the binding constraint on the LunarLander comparison.
Hashed tile coding with linear semi-gradient agents removes it, and a generic
Gymnasium adapter extends the same comparison to other environments. These
require a configured Python via `lunar_setup()` and are not run here.

```{r fa, eval = FALSE}
lunar_setup("/usr/bin/python3")
differentiate_fa("CartPole-v1", n_train = 300)
differentiate_gym("FrozenLake-v1", make_kwargs = list(is_slippery = FALSE))
```

## Reading the results honestly

No single paradigm proves a rule maximizes; each isolates one failure mode of
myopic control. The trap and self-control catch gradient-bandit melioration;
probability matching catches the rate-tracking rule; differential reinforcement
of low rates separates annealed-exploration value rules from persistently
stochastic ones. The matrix is a fingerprint, not a scalar verdict.
